---
title: "postprocess_US-Bar"
output: html_document
date: "2025-04-09"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Before the workshop:
-   Create an account:
    <https://ameriflux-data.lbl.gov/Pages/RequestAccount.aspx>
-   The `user_id` and `user_email` will be used later ...

# load packages
```{r}
# if(!require(devtools)){install.packages("devtools")}
# devtools::install_github("chuhousen/amerifluxr")
library(amerifluxr) # An R programmatic interface for querying, downloading, and handling AmeriFlux data and metadata.
library(REddyProc)  # Package for processing Eddy Covariance flux data (e.g., gap-filling, partitioning)
library(lubridate)  # Provides tools to work with date and time (e.g., parsing, formatting).
library(tidyverse)  # A collection of packages for data manipulation, visualization, and more.
library(dplyr)
library(bigleaf)  # Provides functions to calculate leaf area index and other plant trait estimations.
library(data.table)  # Offers fast and memory-efficient manipulation of large datasets.
library(ggplot2)
```

# working directory
```{r}
# open a temporary directory
tmp_dir <- tempdir()
```

# 01: download AmeriFlux BASE data
-   BASE Publish: After passing the Data QA/QC assessment, AMP formats
    the submitted flux/met data in the FP Standard format, bundles them
    with BADM\*, and versions and publishes the resulting BASE-BADM data
    with a DOI
    (<https://ameriflux.lbl.gov/data/flux-data-products/base-publish/>)
-   AmeriFlux BASE data pipeline:
    <https://doi.org/10.1038/s41597-023-02531-2>
-   R package `amerifluxr`: <https://github.com/chuhousen/amerifluxr>
-   The Bartlett Experimental Forest (44°06′N, 71°3′W, US-Bar) is a
    temperate deciduous forest located in the northeastern United
    States, in the state of New Hampshire where it lies within the Saco
    Ranger District of the White Mountain National Forest.
-   How to download FLUXNET data? ...

```{r}
# check out data coverage for different sites
full.list = amf_data_coverage(data_product = "BASE-BADM", data_policy = "CCBY4.0")
head(full.list)
print(paste("number of sites included in BASE publish is", nrow(full.list)))

# download single site flux/met data
site_name = "US-Bar"
# floc<- amf_download_base(
#     user_id = "my_user", # change to your user name
#     user_email = "my_email@mail.com", # change to your email
#     site_id = site_name,
#     data_product = "BASE-BADM",
#     data_policy = "CCBY4.0",
#     agree_policy = TRUE,
#     intended_use = "other",
#     intended_use_text = "Flux-LSM_workshop",
#     verbose = TRUE,
#     out_dir = tmp_dir
#   )


# If you do not have an account, you can directly start from here
setwd("~/")
floc =  "AMF_US-Bar_BASE-BADM_7-5.zip"
base <- amf_read_base(file = floc,
                      unzip = TRUE,
                      parse_timestamp = TRUE) 

# check data coverage for meteorology
# visualizes the BASE data availability for selected AmeriFlux sites, variables, and years.
#PPFD_IN	Photosynthetic photon flux density, incoming	µmolPhoton m-2 s-1
#TA	Air temperature	deg C
#RH	Relative humidity, range 0-100	%
#VPD	Vapor Pressure Deficit	hPa
#more variable names: https://ameriflux.lbl.gov/data/aboutdata/data-variables/#base
amf_plot_datayear(site_set = site_name, 
                  var_set = c("PPFD_IN", "TA", "RH", "VPD"),
                  # year_set = c(2004:2023),
                  nonfilled_only = TRUE)

# check data coverage for flux data
amf_plot_datayear(site_set = site_name, 
                  var_set = c("FC", "H", "LE"),
                  # year_set = c(2004:2023),
                  nonfilled_only = TRUE)

# uncomment to check data availability by year (all variables)
# amf_plot_datayear(site_set = site_name, nonfilled_only = FALSE)
```

## Questions:
-   How many sites are included in the AmeriFlux BASE publish?
-   Why is the data coverage of flux data lower than that of
    meteorological variables?

# 02: Organise input data
Please note: The naming conventions for variables are different for
different sites;
```{r}
# subset data for two years
base = base[base$YEAR %in% c(2011,2012), ]
# modify time variables
# SP: would be good to explain what the code does here
base$TIMESTAMP=substr(base$TIMESTAMP_START,1,12)
base$year=as.character(substr(base$TIMESTAMP_START,1,4))
base$month=as.character(substr(base$TIMESTAMP_START,5,6))
base$day= as.character(substr(base$TIMESTAMP_START,7,8))
base$hour=as.numeric(as.character(substr(base$TIMESTAMP_START,9,10)))+ c(0.5,1)
base$date=as.Date(with(base, paste(year, month, day, sep="-")), "%Y-%m-%d")
base$doy=yday(base$date)
base$min= as.character(substr(base$TIMESTAMP,11,12))

# format the data to be used as input for REddyProc
base_df = data.frame('Year' = base$year, 'Day' = base$day, 'Hour' = base$hour, 'Date' = as.Date(base$date),
                      'Min' = base$min,  'Month' = base$month, 'DoY' = base$doy,
                      'NEE' = base$FC_1_1_1, 'LE' = base$LE_1_1_1, 'H' = base$H_1_1_1, # flux data
                      'Rg' = ifelse(base$PPFD_IN_PI_F_1_1_1 < 0, 0, PPFD.to.Rg(base$PPFD_IN_PI_F_1_1_1)), 
                      'Tair' = base$TA_PI_F_1_1_1,
                      'Tsoil' = base$TS_PI_F_1_2_1, 'RH' = ifelse(base$RH_PI_F_1_1_1 > 100, 100, base$RH_PI_F_1_1_1),
                      'VPD' = base$VPD_PI_1_1_1, 'Ustar' = base$USTAR_1_1_1,
                      'TIMESTAMP_START' = as.character(base$TIMESTAMP_START),
                      'TIMESTAMP_END' = as.character(base$TIMESTAMP_END),
                      'PPFD_f' = base$PPFD_IN_PI_F_1_1_1)
head(base_df)
```

# 03: intialize EProc
```{r}
?filterLongRuns
#filterLongRuns : Longer runs, i.e. sequences of numerically identical values, in a series of measurements hint to problems during a noisy measurement, e.g. by sensor malfunction due to freezing. This function, replaces such values in such runs with NA to indicate missing values.

EddyData <- filterLongRuns(base_df, "NEE")
EddyData$Year <- as.numeric(EddyData$Year)
EddyData$Hour <- as.numeric(EddyData$Hour)
EddyData$DoY <- as.numeric(EddyData$DoY)
EddyDataWithPosix <- fConvertTimeToPosix(EddyData, 'YDH', Year = 'Year', Day = 'DoY', Hour  = 'Hour') 

head(EddyDataWithPosix$DateTime)
EProc <- sEddyProc$new(site_name, EddyDataWithPosix, c('NEE','Rg','Tair','VPD', 'Ustar', "H", "LE"))
#SP: I don't understand this, is EProc an object with the attributes we defined (NEE, Rg, Tair etc.) ?
#would be good to mention that
class(EProc)
```

# 04: IQR filtering
-   IQR: interquartile range, which is the difference between the 75th
    percentile (Q3) and the 25th percentile (Q1) of a dataset;
-   Marginal distribution sampling (MDS, Reichstein et al., 2005) has
    been adopted as a standard gap-filling method;
-   Standard implementation of MDS relies on photosynthetic photon flux
    density (PPFD) (or, alternatively, global shortwave radiation) as
    the main driver, and air temperature (Tair) and vapor pressure
    deficit (VPD) as additional factors in the look-up table.
-   We first calculate the IQR of the residuals between MDS gap-filled
    values (without u\* filtering) and measured values.
-   Here, MDS gave us an initial “best guess” of what the true flux was
    for each specific half hour. The choice of 6 times the IQR is a
    conservative approach, meaning that only the most extreme values are
    filtered out (some studies have used a more aggressive threshold of
    3 times the IQR). We repeated the filtering process with 6 times of
    IQR twice to ensure sufficient outliers were removed.
-   More information can be found in Liu et al. (2025):
    <https://doi.org/10.1016/j.agrformet.2025.110438>

```{r}
# gapfill meteorological variables using MDS
EProc$sMDSGapFill('Tair', FillAll = FALSE,  minNWarnRunLength = NA)
EProc$sMDSGapFill('Rg', FillAll = FALSE,  minNWarnRunLength = NA)
EProc$sMDSGapFill('VPD', FillAll = FALSE,  minNWarnRunLength = NA)
# use MDS to get the "best guess" of true flux
EProc$sMDSGapFill('NEE') 

# Calculate residuals and identify outliers
residual <- EProc$sTEMP$NEE_orig - EProc$sTEMP$NEE_fall
IQR <- IQR(residual, na.rm = TRUE)
outlier <- ifelse(abs(residual) > (IQR * 6), 1, 0)
EddieC <- data.frame(
    sDateTime = EProc$sTEMP$sDateTime,
    NEE_orig = EProc$sTEMP$NEE_orig,
    Ustar = EProc$sDATA$Ustar,
    NEE_fall = EProc$sTEMP$NEE_fall,
    residual = residual,
    outlier = outlier
  )
  
# Rename columns
colnames(EddieC) <- c('sDateTime', 'NEE_orig', 'Ustar', 'NEE_fall', 'residual', 'outlier')
  
# Filter out outliers
EddieC$NEE_filt <- dplyr::if_else(EddieC$outlier > 0, NA_real_, EddieC$NEE_orig)
EddieC$year <- substr(EddieC$sDateTime, 1, 4)
EddieC$doy <- strftime(EddieC$sDateTime, format = "%j")
# Plot the initial outlier detection
EddieC %>%
    arrange(as.factor(outlier)) %>%
    ggplot(aes(y = NEE_orig, x = as.numeric(doy), color = as.factor(outlier))) +
    geom_point(shape = 20, alpha = 0.4) +
    theme_minimal() +
    labs(x = 'Day of year', y = 'NEE') +
    scale_color_manual(values = c('skyblue', 'red')) +
    facet_wrap(~year) + 
    ylim(c(-50, 50)) +
    ggtitle("intial outlier detection")
  
# Re-run the outlier test after initial filtering
EddieC$residual2 <- EddieC$NEE_filt - EddieC$NEE_fall
EddieC$IQR2 <- IQR(EddieC$residual2, na.rm = TRUE)
EddieC$outlier2 <- ifelse(abs(EddieC$residual2) > EddieC$IQR2 * 6, 1, 0)
EddieC$NEE_filt2 <- ifelse(EddieC$outlier2 == 0, EddieC$NEE_filt, NA)
  
# Plot the re-run outlier detection
EddieC %>%
    arrange(as.factor(outlier2)) %>%
    ggplot(aes(y = NEE_orig, x = as.numeric(doy), color = as.factor(outlier2))) +
    geom_point(shape = 20) +
    theme_minimal() +
    labs(x = 'Day of year', y = 'NEE') +
    scale_color_manual(values = c('skyblue', 'red')) +
    facet_wrap(~year) + 
    ylim(c(-50, 50)) +
    ggtitle("re-run outlier detection")

# Remove outliers from the main data
EProc$sDATA$NEE <- EddieC$NEE_filt2
```

# 05: u\* filtering
-   u\*: Friction velocity (m/s) is a measure of turbulent mixing in the
    lower atmosphere near the surface
-   It can be thought of as a velocity scale, a representative value for a 'turbulent velocity'
-   u\* is defined using the vertical flux of horizontal momentum
-   (u\*)\^2 = overbar(u'w')\^2 + overbar(v'w')\^2
-   CO2 transport is heavily influenced by turbulent mixing.
-   Unfavorable conditions could be detected by inspecting the relationship of nighttime NEE vs. u\*.

Why is u\* filtering needed? 

- To filter out period with low turbulence
- For e.g., at night when winds are calm, the atmosphere often becomes stable with low turbulence and low mixing locally.
  - This could lead to CO2 pooling near the surface or within canopies.
- EC towers may underestimate respiration fluxes as CO2 is not being transported upward efficiently to be measured. 

Different treatments of u\* thresholds; - At low u\* values, a
negatively biased respiration is measured; - u\* threshold is the
minimum u\* above which respiration reachers a plateau; (check out the
figure under /plots) - user-specific u\* threshold - Single u\*
threshold - annual varying u *- seasonal varing u* - read more:
<https://cran.r-project.org/web/packages/REddyProc/vignettes/uStarCases.html>

```{r}
# EProc$sPlotNEEVersusUStarForSeason(format = "pdf") # keep this line commented, please check the figure under /plots
set.seed(2000)
uStarTh <- EProc$sEstUstarThresholdDistribution(nSample = 10L, probs = c(0.05, 0.5, 0.95)) # use nSample = 1000L for true analysis
print(uStarTh)

# Define aggregation mode
EProc$sGetUstarScenarios() # by default, annual varing u* is used
# EProc$useSeaonsalUStarThresholds()# seasonal varing u*
# EProc$useAnnualUStarThresholds() # annual varing u*
  
# Here, we use the single u* threshold
uStar <- uStarTh  %>%
  dplyr::filter( aggregationMode == "single") %>%
  dplyr::select( "uStar", "5%", "50%", "95%")
uStarDf <- cbind(season=na.omit(unique(uStarTh$season)), uStar)
EProc$sSetUstarScenarios(uStarDf)
EProc$sGetUstarScenarios()
```

Question: What aggregationMode should I use for my site?

# 06: MDS: gapfil NEE

-   The MDS combines two gap-filling techniques: the “look-up” table
    (LUT) and the “mean diurnal course” (MDC). \## “look-up” table (LUT)
    Fluxes are expected similar if they are

-   at similar environmental conditions

    -   Rg $\pm$ 50 $W m^{-2}$, Tair $\pm$ 2.5 $°C$, and VPD $\pm$ 5.0
        $h Pa$

-   and close in time

    -   increasing time window until enough observations \## “mean
        diurnal course” (MDC) Fluxes are expected similar if they are

-   at the same time of the day

-   not too many days away

-   The methods are effective for short gaps as the missing values are
    replaced by the average of response variables under similar weather
    conditions in a small-time window; Combines LUT and MDC

Quality flag increases with fewer variables and larger time windows

-   0 Observations
-   1 gap-filled with good quality
-   $>1$ gap-filled with lower quality

```{r}
# fingerplot: inspect gaps needed to be filled
EProc$sPlotFingerprintY("NEE", Year = 2011); EProc$sPlotFingerprintY("NEE", Year = 2012)
# use MDS to gapfill flux data
EProc$sMDSGapFillUStarScens('NEE')

# gaps in flux data are filled 
EProc$sPlotFingerprintY("NEE_50._f", Year = 2011); EProc$sPlotFingerprintY("NEE_50._f", Year = 2012)
```

Questions: - Can I use more than 3 drivers to gapfill flux data? -\

# 07: partitioning NEE into Reco and GPP
-   NEE = Reco - GPP
-   Several methods were developed to partition NEE into Reco and GPP;
-   please note: Reco is modeled and GPP is computed as Reco - NEE;
-   night-time partitioning:
-   day-time partitioning:
-   modified daytime partitioning:

```{r}
EProc$sSetLocationInfo(LatDeg = 44.0646, LongDeg = -71.2881, TimeZoneHour = -5)
# TimeZoneHour: time zone offset from UTC (Coordinated Universal Time), without daytime saving

EProc$sFillVPDFromDew() # fill longer gaps still present in VPD_f
EProc$sMRFluxPartitionUStarScens("NEE") # night-time
# EProc$sGLFluxPartitionUStarScens() # day-time
# EProc$sTKFluxPartitionUStarScens() # modified day-time
names(EProc$sTEMP)
```

# 08: final output
Question: What is the difference between FCO2 and NEE?:
<https://ameriflux.lbl.gov/data/aboutdata/data-variables/>

```{r}
FilledEddyData <- EProc$sExportResults()
combined.df <- cbind(EddyData, FilledEddyData)
names(combined.df)

#Have a look at the gapfilled data

# plot diurnal pattern of FC
ggplot(combined.df, aes(x = Hour, y = NEE_50._f)) +
  geom_point(col = "grey") +  
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  stat_summary(fun = mean, geom = "line") + 
  facet_wrap(~ interaction(Month, Year)) +
  theme_minimal() +
  ggtitle("FC") +  ylab(expression(FCO[2]~'('*gC~m^{-2}~y^{-1}*')'))
 

# plot diurnal pattern of GPP
ggplot(combined.df, aes(x = Hour, y = `GPP_50%_f`)) +
  geom_point(col = "grey") +  
  stat_summary(fun = mean, geom = "line") + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +  
  facet_wrap(~ interaction(Month, Year)) +
  theme_minimal() +
  ggtitle("GPP") + ylab(expression(GPP~'('*gC~m^{-2}~y^{-1}*')'))

# plot diural pattern of Reco
ggplot(combined.df, aes(x = Hour, y = `Reco_50%`)) +
  geom_point(col = "grey") +  
  stat_summary(fun = mean, geom = "line") + 
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +  
  facet_wrap(~ interaction(Month, Year)) +
  theme_minimal() +
  ggtitle("Reco") + ylab(expression(Reco~'('*gC~m^{-2}~y^{-1}*')'))
```
# To do: 


# References
-   Wutzler, Thomas, et al. "Basic and extensible post-processing of
    eddy covariance flux data with REddyProc." Biogeosciences 15.16
    (2018): 5015-5030.
-   Reichstein, Markus, et al. "On the separation of net ecosystem
    exchange into assimilation and ecosystem respiration: review and
    improved algorithm." Global change biology 11.9 (2005): 1424-1439.
-   Liu, Yujie, et al. "Robust filling of extra-long gaps in eddy
    covariance CO2 flux measurements from a temperate deciduous forest
    using eXtreme Gradient Boosting." Agricultural and Forest
    Meteorology 364 (2025): 110438.
